Shaping strategies and standardisation of faceted spheroids from Qianshangying (North China)

### 2025-06-19 analysis/paper/paper.qmd
updates

```{r}
# 读取表格
library(tidyverse)
library(readxl)
library(here)
library(stringr)

@@

Spherical Harmonic (SPHARM) analysis is a method for describing 3D shapes using a series of mathematical functions defined on the surface of a sphere. It works like a 3D version of Fourier analysis, breaking down complex shapes into components of different spatial frequencies. Each shape is first mapped onto a unit sphere and then expressed as a combination of spherical harmonic functions $Y_l^m(\theta, \phi)$. The shape is represented by a set of coefficients $\hat{f}(l, m)$, which capture how much each function contributes. Here, θ and ϕ are angles in spherical coordinates, l and m are order and degree respectively. Lower-order terms describe the general shape, while higher-order terms capture finer details.

@@

This study represents the first attempt in archaeology to apply SPHARM for the classification of spheroids and cores. Specifically, we included a set of standard geometric models—such as sphere, rounded cube, ellipsoid (1.2:1:0.8) rounded rectangular prism (1.2:1:1), and disc-like (flattened, and like discoid core)—in the SPHARM analysis and UMAP. Spheroids and cores were positioned in this position based on their morphological similarity to these reference models: individuals that cluster more closely to a specific standard form exhibit corresponding shape characteristics.

@@

```{r}

@@

```{r}
# Pearson correlation test

@@

p_value_pearson_sci <- pretty_print_sci(p_value)
```

```{r}
# Mean SHE

@@

```{r}
# Kruskal–Wallis test2
# on SHE among 4 core types

@@

cv_she_sub_spheroids
cv_she_poly_multi
```

@@

We conducted shape analysis based on spherical harmonic decomposition, setting the maximum degree to 20 (lmax=20). This choice ensures that sufficient geometric detail is captured while avoiding high-frequency noise and overfitting artifacts such as the Gibbs phenomenon. After normalization, all models have their degree 0 (l=0) coefficient set to 1, allowing for meaningful comparison across different models.

@@

To visualize the structure of shape variation, we applied UMAP to the normalized power spectrum. The UMAP projection reveals distinct clusters in the first two dimensions of the shape space. These clusters correspond to reference models, sphere, rounded cube, ellipsoid, rounded rectangular prism and disc-shaped core. This demonstrates that these cores occupy distinguishable regions in the harmonic shape space, with proximity to the sphere or elongation/flattening along specific axes emerging as key differentiating features. Most artifacts typologically classified as spheroids or sub-spheroids were grouped into the “rounded cube” and “sphere” clusters, indicating their consistent, near-spherical shapes. However, some Multifacial cores and Polyhedron also appeared in the near-spherical region. This can be attributed to their high cortex coverage or limited flake removals, which likely preserved their originally rounded forms and led to their grouping alongside more intentionally shaped spheroids.

We then tested whether the UMAP-based groupings also correspond to differences in sphericity using a Kruskal–Wallis test, which yielded $\chi^2$ = `r round(H_value, 2)`, df = `r df_kruskal`, p = $`r p_value_kruskal_sci`$. The result shows statistically significant differences in sphericity among the shape clusters, further supporting the effectiveness of the spherical harmonic approach in capturing meaningful morphological variation.

@@

The diacritical analysis of spheroid surfaces reveals a structured and recurrent reduction sequence comprising the following phases. Phase 1 was a unidirectional or bidirectional total exploitation, belonging to one kind of volume exploitation. Large and flat surfaces of the cobble were selected as platforms for striking toward the opposite surface. Knapping proceeded until the peripheral volume was fully exploited, resulting in an approximately cubic form. Phase 2 was peripheral centripetal exploitation. In this phase, previous removals in phase 1 serve as platforms for centripetal flaking around the core periphery. This represents a surface exploitation strategy. A notable feature is the angle between the face's ventral surface and platform (i.e. core edge angle), which often exceeds 90°, forming a secant plan [citation] rather than parallel planes. This phase contributes significantly to the increasingly spherical shape of the artefact. Phase 3 was the production of small scars on ridges or intersections of previous flake scars, resulting in a rounding and smoother surface, further enhancing sphericality. It is important to note that due to the high degree of reduction intensity observed in most spheroids, phases prior to the visible sequence were often obliterated. We hypothesize the existence of a Phase 0, potentially involving roughing-out of the cobble or an earlier cycle of Phases 1–3.

@@

<!-- Based on this diacritical analysis and scar orientation analysis, we can summarize the technical strategy employed in the production of faceted spheroids in QSY. This strategy consists of a series of sequential phases: a potential rough-out phase, volume exploitation aimed at forming a cubic shape, surface exploitation to approximate a spheroidal geometry, edge refinement, and potentially the repetition of earlier phases. The existence of a repetitive phase is supported by the observation that while the original cobble blanks in QSY display considerable size variation, the finished spheroids—although some are relatively large—mostly cluster around a diameter of 8–10 cm. This suggests that the production of spheroids may have involved multiple cycles of reduction, and further implies that a diameter of 8–10 cm was considered an optimal or intended size in the knapping process. BM: maybe move to discussion -->

@@

knitr::kable(data_raw2, caption = "Summary of core typology and morphological parameters.")

@@

# 打印结果查看
# print("Spearman correlation matrix:")
# print(round(cor_matrix, 4))
# print("P-value matrix:")
# print(round(p_matrix, 4))

@@

Among the shape descriptors of spheroids, several significant correlations were observed. Mean edge angle was positively correlated with sphericity (ρ = `r cor_angle_sph`, p = $`r p_angle_sph`$) and negatively correlated with SPHARM power (ρ = `r cor_angle_she`, p = $`r p_angle_she`$), indicating that greater edge angles are associated with more spherical and regular forms. As noted in previous studies, mean edge angle is a key variable in capturing the morphology of spheroids [citation]. This correlation between edge angle and sphericity may help explain why spheroids were often interpreted as exhausted cores in earlier research [citation]. Indeed, spheroids must be significantly reduced, as achieving a near-perfect spherical shape typically results in increased edge angles, making further flake removal difficult. However, the reverse is not necessarily true—not all exhausted cores evolve into spheroids.

>>>>

```{r, echo=FALSE}
# 读取表格
library(tidyverse)
library(readxl)
library(here)
library(stringr)

@@

Spherical Harmonic (SPHARM) analysis is a method for describing 3D shapes using a series of mathematical functions defined on the surface of a sphere. It works like a 3D version of Fourier analysis, breaking down complex shapes into components of different spatial frequencies. Each shape is first mapped onto a unit sphere and then expressed as a combination of spherical harmonic functions $Y_l^m(\theta, \phi)$. The shape is represented by a set of coefficients $\hat{f}(l, m)$, which capture how much each function contributes. Here, θ and ϕ are angles in spherical coordinates, l and m are order and degree respectively. Lower-order terms describe the general shape, while higher-order terms capture finer details (@fig-SPHARM-Reconstruction-by-Degree).

@@

This study represents the first attempt in archaeology to apply SPHARM for the classification of spheroids and cores. Specifically, we included a set of standard geometric models—such as sphere, rounded cube, ellipsoid (1.2:1:0.8) and box (1.2:1:1), and disc-like (flattened, and like discoid core)—in the SPHARM analysis and UMAP. Spheroids and cores were positioned in this position based on their morphological similarity to these reference models: individuals that cluster more closely to a specific standard form exhibit corresponding shape characteristics.

@@

```{r, echo=FALSE}

@@

```{r, include=FALSE}
# Pearson correlation test

@@

p_value_pearson_sci <- pretty_print_sci(p_value)

# 画图
ggplot(data_clean, aes(x = Sphericity, y = `SPHARM power`)) +
  # 散点（深红色）
  geom_point(size = 3, alpha = 0.9, color = "#B22222") +

  # 拟合线（深蓝灰）
  geom_smooth(method = "lm", se = TRUE, color = "#1C3041", fill = "#A3B1C6", linetype = "solid") +

   scale_x_continuous(breaks = seq(0.4, 1.0, by = 0.04)) +

  # 自定义配色主题
  theme_minimal(base_size = 15) +
  theme(
    panel.grid.major = element_line(color = "#DDDDDD"),
    panel.grid.minor = element_blank(),
    axis.title = element_text(color = "#333333"),
    axis.text = element_text(color = "#333333")
  ) +

  # 标题与坐标轴标签
  labs(
    x = "Sphericity",
    y = "SPHARM Power"
  ) +
  annotate("text",
         x = Inf, y = Inf,
         hjust = 1.1, vjust = 1.2,
         label = paste0("r = ", round(correlation, 3),
                        "\n", "p = ", pretty_print_sci(p_value),
                        "\n", "n = ", nrow(data_clean)),
         size = 5, color = "#1C3041")
```

```{r, echo=FALSE}
# Mean SHE

@@

```{r, echo=FALSE}
# Kruskal–Wallis test2
# on SHE among 4 core types

@@

```{r setup_umap_data, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggrepel)
library(here)

# Load data
umap_data <- read_csv(here("analysis/data/derived_data/umap_lmax20.csv"))

# Lowercase category and set factor levels
umap_data <- umap_data %>%
  mutate(category_lower = tolower(category)) %>%
  mutate(category_lower = fct_relevel(category_lower,
    "polyhedron", "multifacial", "idealmodel", "subspheroid", "spheroid"
  ))

# Set point size (highlight idealmodel)
umap_data <- umap_data %>%
  mutate(point_size = if_else(category_lower == "idealmodel", 4.5, 3))

# KMeans clustering (k=4)
set.seed(123)
kmeans_result <- kmeans(umap_data[, c("x", "y")], centers = 4)

# Add clustering results
umap_data <- umap_data %>%
  mutate(
    cluster = factor(kmeans_result$cluster),
    is_idealmodel = if_else(category_lower == "idealmodel", "yes", "no")
  )

# Save x/y limits for consistent plotting
x_limits <- range(umap_data$x)
y_limits <- range(umap_data$y)

expand_range <- function(lim, ratio) {
  range_width <- diff(lim)
  return(c(lim[1] - range_width * ratio, lim[2] + range_width * ratio))
}

# 原始坐标范围
x_limits <- range(umap_data$x)
y_limits <- range(umap_data$y)

# 分别扩展 x 和 y
x_limits_expanded <- expand_range(x_limits, 0.10)  # 横轴扩展 10%
y_limits_expanded <- expand_range(y_limits, 0.15)  # 纵轴扩展 15%

# Subset for labeling
label_data <- umap_data %>%
  filter(category_lower %in% c("idealmodel", "spheroid", "subspheroid"))
```

```{r umap_by_category, include=FALSE}
category_colors <- c(
  "idealmodel" = "black",
  "spheroid" = "red",
  "subspheroid" = "orange",
  "multifacial" = "lightgray",
  "polyhedron" = "gray40"
)

ggplot(umap_data, aes(x = x, y = y)) +
  geom_point(aes(color = category_lower, size = point_size), alpha = 0.9, shape = 16) +
  #geom_text_repel(
    #data = label_data,
    #aes(label = filename),
    #size = 3,
    #max.overlaps = 100,
    #show.legend = FALSE
  #) +
  scale_color_manual(values = category_colors, name = "Category") +
  scale_size_identity() +
  coord_cartesian(xlim = x_limits_expanded, ylim = y_limits_expanded)+ #确保范围一致
  theme_minimal(base_size = 14) +
  labs(
    x = "UMAP 1",
    y = "UMAP 2"
  ) +
  theme(
    legend.position = "non",
    # plot.title = element_text(face = "bold")
  )
```

```{r umap_by_cluster, include=FALSE}
cluster_colors <- RColorBrewer::brewer.pal(4, "Set2")

ggplot(umap_data, aes(x = x, y = y)) +
  geom_point(
    data = filter(umap_data, is_idealmodel == "no"),
    aes(color = cluster),
    alpha = 0.8,
    size = 2.5
  ) +
  geom_point(
    data = filter(umap_data, is_idealmodel == "yes"),
    aes(color = cluster),
    size = 3.5,
    show.legend = FALSE
  ) +
  stat_ellipse(
    aes(color = cluster),
    type = "norm",
    level = 0.95,
    linetype = "dashed",
    show.legend = FALSE
  ) +
  scale_color_manual(
    values = cluster_colors,
    name = "Cluster"
  ) +
  coord_cartesian(xlim = x_limits_expanded, ylim = y_limits_expanded) +  # 保证坐标一致
  theme_minimal(base_size = 14) +
  theme(legend.position = "none") +
  labs(
    x = "UMAP 1",
    y = "UMAP 2",
    #caption = "Dashed ellipses show 95% confidence intervals"
  )
```

```{r, echo=FALSE}
#| label: fig-SPHARM-Reconstruction-by-Degree
#| fig-cap: "Shape reconstruction of archaeological spheroids using SPHARM (various lmax levels)"

library(here)
knitr::include_graphics(here("analysis/paper/figures/SPHARM_reconstruction.png"))

```{r, echo=FALSE}
#| label: fig-UMAP-SPHARM-clusters
#| fig-cap: "UMAP plot showing distinct clusters of shapes derived from SPHARM"

library(here)
knitr::include_graphics(here("analysis/paper/figures/UMAP_SPHARM_clusters.png"))

```{r, echo=FALSE}
#| label: fig-UMAP-SPHARM-samples
#| fig-cap: "UMAP plot of SPHARM-derived shape features colored by types"

library(here)
knitr::include_graphics(here("analysis/paper/figures/UMAP_SPHARM_samples.png"))

```{r, echo=FALSE}
#| label: fig-Correlation-SPHARMpower-Sphericity
#| fig-cap: "Correlation analysis between SPHARMpower and Sphericity"

library(here)
knitr::include_graphics(here("analysis/paper/figures/Correlation_SPHARMpower_Sphericity.png"))

@@

We conducted shape analysis based on spherical harmonic decomposition, setting the maximum degree to 20 (lmax=20,@fig-SPHARM-Reconstruction-by-Degree). This choice ensures that sufficient geometric detail is captured while avoiding high-frequency noise and overfitting artifacts such as the Gibbs phenomenon. After normalization, all models have their degree 0 (l=0) coefficient set to 1, allowing for meaningful comparison across different models.

@@

To visualize the structure of shape variation, we applied UMAP to the normalized power spectrum. The UMAP projection reveals distinct clusters in the first two dimensions of the shape space. These clusters correspond to reference models, sphere, rounded cube, ellipsoid, box and disc-shaped core (@fig-UMAP-SPHARM-clusters). This demonstrates that these cores occupy distinguishable regions in the harmonic shape space, with proximity to the sphere or elongation/flattening along specific axes emerging as key differentiating features. Most artifacts typologically classified as spheroids or sub-spheroids were grouped into the “sphere” , “cube” and "ellipsoid" clusters, indicating their consistent, near-spherical shapes (@fig-UMAP-SPHARM-samples). However, some Multifacial cores and Polyhedron also appeared in the near-spherical region. This can be attributed to their high cortex coverage or limited flake removals, which likely preserved their originally rounded forms and led to their grouping alongside more intentionally shaped spheroids.

We then tested whether the UMAP-based groupings also correspond to differences in sphericity using a Kruskal–Wallis test, which yielded $\chi^2$ = `r round(H_value, 2)`, df = `r df_kruskal`, p = $`r p_value_kruskal_sci`$. The result shows statistically significant differences in sphericity among the shape clusters, further supporting the effectiveness of the spherical harmonic approach in capturing meaningful morphological variation (@fig-Correlation-SPHARMpower-Sphericity).

@@

The diacritical analysis of spheroid surfaces reveals a structured and recurrent reduction sequence comprising the following phases (@fig-spheroids-2797-1098). Phase 1 was a unidirectional or bidirectional total exploitation, belonging to one kind of volume exploitation. Large and flat surfaces of the cobble were selected as platforms for striking toward the opposite surface. Knapping proceeded until the peripheral volume was fully exploited, resulting in an approximately cubic form. Phase 2 was peripheral centripetal exploitation. In this phase, previous removals in phase 1 serve as platforms for centripetal flaking around the core periphery. This represents a surface exploitation strategy. A notable feature is the angle between the face's ventral surface and platform (i.e. core edge angle), which often exceeds 90°, forming a secant plan \[citation\] rather than parallel plan. This phase contributes significantly to the increasingly spherical shape of the artefact. Phase 3 was the production of small scars on ridges or intersections of previous flake scars, resulting in a rounding and smoother surface, further enhancing sphericality. It is important to note that due to the high degree of reduction intensity observed in most spheroids, phases prior to the visible sequence were often obliterated. We hypothesize the existence of a Phase 0, potentially involving roughing-out of the cobble or an earlier cycle of Phases 1–3.

@@

```{r, echo=FALSE}
#| label: fig-spheroids-2797-1098
#| fig-cap: "Diacritical analysis of QSY spheroids"

library(here)
knitr::include_graphics(here("analysis/paper/figures/spheroids_2797_1098.png"))

```{r, echo=FALSE}
#| label: fig-shaping-process
#| fig-cap: "hypothesis of QSY spheroid shaping process"

library(here)
knitr::include_graphics(here("analysis/paper/figures/shaping_process.png"))

Based on this diacritical analysis and scar orientation analysis, we can summarize the technical strategy employed in the production of faceted spheroids in QSY. This strategy consists of a series of sequential phases (@fig-shaping-process): a potential rough-out phase, volume exploitation aimed at forming a cubic shape, surface exploitation to approximate a spheroidal geometry, edge refinement, and potentially the repetition of earlier phases. The existence of a repetitive phase is supported by the observation that while the original cobble blanks in QSY display considerable size variation, the finished spheroids—although some are relatively large—mostly cluster around a diameter of 8–10 cm. This suggests that the production of spheroids may have involved multiple cycles of reduction, and further implies that a diameter of 8–10 cm was considered an optimal or intended size in the knapping process. <!-- BM: maybe move to discussion -->

@@

knitr::kable(data_raw2, caption = "Summary of spheroid shaping intensity and shape features.")

@@

@@

Among the shape descriptors of spheroids, several significant correlations were observed. Mean edge angle was positively correlated with sphericity (ρ = `r cor_angle_sph`, p = $`r p_angle_sph`$) and negatively correlated with SPHARM power (ρ = `r cor_angle_she`, p = $`r p_angle_she`$), indicating that greater edge angles are associated with more spherical and regular forms. As noted in previous studies, mean edge angle is a key variable in capturing the morphology of spheroids \[citation\]. This correlation between edge angle and sphericity may help explain why spheroids were often interpreted as exhausted cores in earlier research \[citation\]. Indeed, spheroids must be significantly reduced, as achieving a near-perfect spherical shape typically results in increased edge angles, making further flake removal difficult. However, the reverse is not necessarily true—not all exhausted cores evolve into spheroids.
